\documentclass[12pt]{article}


\usepackage[dvips,letterpaper,margin=0.75in,bottom=0.75in]{geometry}
\usepackage{cite}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{amsmath}

\usepackage[american,fulldiode]{circuitikz}

\begin{document}
\ctikzset{bipoles/thickness=1}
\ctikzset{bipoles/length=.6cm}

\title{Circuit Design}

\maketitle


\section{Design Tasks}
\begin{enumerate}
\item Choose an LED from Digikey.  Sketch a simple circuit that lights
  the LED using a DC voltage source and resistor.  Choose a DC voltage
  and a resistor value which ensures that the LED is lit brightly.

\item Using a single generic NPN bipolar transistor (assume
  $\beta=100$), design a follower that operates between 0 and 16
  volts.  Use a voltage divider at the input and an emitter resistor
  to set the DC operating point such that quiescent bias current is
  100 mA and the AC output has the maximum dynamic range.  You may
  assume all capacitors are infinitely large (short circuits to AC)
  but specify all resistor values.  Calculate the input impedance of
  your circuit.

\item Suppose that input impedance of your circuit is insufficent.
  Using the same single transistor, loosen one of the other design
  constraints in order to achieve a factor of ten higher input
  impedance.

\item Using a single generic NPN bipoloar transistor (assume
  $\beta=100$), design a common emmitter amplifier that operates
  between 0 and 10 volts.  Use a voltage divider at the input and an
  emitter and collector resistor.  Set the DC operating point such
  that quiescent bias current is 100 mA and the AC output has the
  maximum dynamic range.  Set the gain to five.  You may assume all
  capacitors are infinitely large (short circuits to AC) but specify
  all resistor values.

\item Choose an operational amplifier on Digikey and design a circuit
  with a gain of 50 up to 10 MHz using a single stage.
  
\item Using a single generic operational amplifier, design an active
  low-pass filter with a corner frequency of 10 MHz and an input
  impedance of $1~\rm M\Omega$ at 10~MHz.

\item For this design, use (multiple instances) of a generic
  operational amplifier.  Assume that is has a gain-bandwidth product
  of $100~\rm MHz$.  Design a circuit that has input impedance of
  $10~\rm M\Omega$ and a gain of 1000 for frequencies up to
  $10~\rm~MHz$

\item Lookup the specs for the LT3045 linear voltage regulator.  Sketch a circut which configures the LT3045 to deliver a fixed 10 V regulated output.
  
  
\end{enumerate}



 
\end{document}
